---
title: "Correlation Heatmap Visual"
output: html_document
date: "2025-12-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Correlation Heatmap

This visualization shows the correlation between funding variables and outcome/operational metrics.

**Required files:**
- `data/02-processed/startup_data.csv` - Startup funding, status, and metrics data
- `data/02-processed/dataset3_cleaned.csv` - Startup failure data

```{r heatmap, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
library(plotly)
library(readr)

# Load the datasets
startup_data <- read_csv("data/02-processed/startup_data.csv", show_col_types = FALSE)
failures_all <- read_csv("data/02-processed/dataset3_cleaned.csv", show_col_types = FALSE)

# Join with failures data
df <- startup_data %>%
  left_join(failures_all, by = c("name" = "Name")) %>%
  mutate(
    failure_amount = as.numeric(gsub("[^0-9.]", "", How.Much.They.Raised)),
    total_amount_raised = coalesce(funding_total_usd, failure_amount),
    years_of_operation = parse_number(as.character(Years.of.Operation)),
    # Create status_binary (1 = acquired, 0 = closed)
    status_binary = as.numeric(status == "acquired")
  )

# Define variables for correlation
funding_vars <- c(
  "has_roundA",
  "has_roundB",
  "has_roundC",
  "has_roundD",
  "has_angel",
  "has_VC",
  "total_amount_raised"
)

outcome_vars <- c(
  "status_binary",
  "relationships",
  "milestones",
  "age_last_funding_year",
  "age_last_milestone_year",
  "years_of_operation"
)

# Filter to only include variables that exist in the dataframe
funding_vars <- funding_vars[funding_vars %in% names(df)]
outcome_vars <- outcome_vars[outcome_vars %in% names(df)]

# Calculate correlation matrix
corr_mat <- cor(
  df[, funding_vars],
  df[, outcome_vars],
  use = "pairwise.complete.obs"
)

# Clean names for display
clean_names <- list(
  has_roundA = "Series A",
  has_roundB = "Series B",
  has_roundC = "Series C",
  has_roundD = "Series D",
  has_angel = "Has Angel",
  has_VC = "Has VC",
  status_binary = "Status (Acquired)",
  relationships = "# Relationships",
  milestones = "# Milestones",
  age_last_funding_year = "Age at Last Funding",
  age_last_milestone_year = "Age at Last Milestone",
  years_of_operation = "Years of Operation",
  total_amount_raised = "Total Raised (USD)"
)

hover_text <- round(corr_mat, 2)

x_labels <- unlist(clean_names[colnames(corr_mat)])
y_labels <- unlist(clean_names[rownames(corr_mat)])

# Create the heatmap
p <- plot_ly(
  x = x_labels,
  y = y_labels,
  z = corr_mat,
  type = "heatmap",
  colorscale = list(
    list(0, "blue"),
    list(0.5, "white"),
    list(1, "red")
  ),
  zmin = -1,
  zmax = 1,
  text = hover_text,
  hovertemplate = paste(
    "Funding: %{y}<br>",
    "Outcome: %{x}<br>",
    "Correlation: %{text}<extra></extra>"
  )
) %>%
  layout(
    title = list(
      text = "<b>Correlation between Funding, Outcomes, and Failure Metrics</b>",
      x = 0.5,
      font = list(size = 22)
    ),
    xaxis = list(
      title = "<b>Outcomes/Operational Metrics</b>",
      tickangle = 45,
      tickfont = list(size = 14)
    ),
    yaxis = list(
      title = "<b>Funding Variables</b>",
      tickfont = list(size = 14)
    ),
    margin = list(t = 120, l = 140, b = 140)
  )

p

# Print correlation matrix summary
cat("\nCorrelation Matrix Summary:\n")
cat("Funding variables:", paste(funding_vars, collapse = ", "), "\n")
cat("Outcome variables:", paste(outcome_vars, collapse = ", "), "\n")
```

## How to Run This Visualization

1. Open this file in RStudio
2. Ensure the following files are in your working directory:
   - `data/02-processed/startup_data.csv`
   - `data/02-processed/dataset3_cleaned.csv`
3. Install required packages: `install.packages(c("dplyr", "stringr", "plotly", "readr"))`
4. Click "Knit" or run all code chunks to generate the visualization
